Semantic Linearity in a Lazy Optimising Compiler

### 2024-02-12 icfp.tex
RM edits recursive lets and cases
Missing: case of variable

%\vspace{-.5cm}
\paragraph{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}
% Second, we look into recursive let bindings.
For the most part,
recursive let bindings behave as non-recursive let bindings, i.e. we must use them \emph{at
most once} because, when evaluated, the linear resources used in the binders
bodies are consumed. The defining property of a group of mutually recursive let
bindings is that the binders bound in that group can occur, without restrictions, in
the bodies of those same binders. % The same way that, in a let body, evaluating
% a binding that uses some resource exactly once consumes that resource once,
% using the binding in its own definition also entails using that resource once.
Consider the following program, that calls a recursive let-bound function
defined in terms of the linear resource \incode{x} and itself:
%
\begin{notyet}
\begin{lstlisting}
f3 :: Bool -> a *' $\lolli$ '* a
f3 bool x = let go b = case b of
                     True -> x
                     False -> go (not b)
                   in go bool
\end{lstlisting}
\end{notyet}
%
Function \incode{f3} is semantically linear because, iff it is consumed exactly once,
then \incode{x} is consumed exactly once. We can see this by case analysis on \incode{go}'s argument:
%\begin{itemize}%
%\item
when \incode{bool} is \incode{True}, we'll use the resource \incode{x};
%\item
 when \incode{bool} is \incode{False}, we recurse by calling \incode{go} on \incode{True}, which in turn will use the resource \incode{x}.
%\end{itemize}
In \incode{go}'s body, \incode{x} is used directly in one branch and indirectly in the
other, by recursively calling \incode{go} (which we know will result in using \incode{x}
linearly).
% ROMES:TODO:!!
%\todo{this bit is quite hard to explain. It is some sort of cyclic
%argument -- we kind of assume go uses x linearly s.t. when go itself is used
%then we're using x linearly. recursion...}
%

@@

\paragraph{Case expressions}
Finally, we discuss case expressions,
which are the key ingredient that brings together the
semantic linearity insights developed thus far given that
\emph{case expressions drive evaluation} and semantic linearity can only be
understood in terms of how expressions are evaluated.

Up until now, there
have been no examples in which linear resources are \emph{fully consumed} in
the bodies of linear functions. In other words, all example functions so far
return a value that has to be itself consumed exactly once to ensure the linear
argument is, in turn, consumed exactly once -- as opposed to functions whose
application simply needs to be evaluated to guarantee its linear argument is
consumed (e.g.,~functions that return an unrestricted value).
%
%The latter are of particular relevance because linearly-typed abstractions
%usually require such a function to provide newly-created linear resources to.
%
% For example, the entry point to the linear array API presented in Linear
% Haskell takes such a function as its second argument:
% \begin{working}
% \begin{lstlisting}
% newMArray :: Int -> (MArray a *' $\lolli$ '* Unrestricted b) *' $\lolli$ '* b
% \end{lstlisting}
% \end{working}
%(The second argument is a function that consumes |MArray a| linearly
%and returns an unrestricted result -- we don't need to consume said result
%exactly once to guarantee that |MArray a| is used linearly in the function
%body.)
%
% ROMES:TODO: Há qualquer coisa que não gosto nada! na coerência deste paragrafo
%
Case expressions enable us to consume resources and so write
functions that fully consume their linear arguments.
% CORTEI
To understand exactly how,
we turn to the definition of \emph{consuming} a resource from Linear
Haskell~\cite{cite:linearhaskell}:
%begin{itemize}
% \item
to consume a value of atomic base type (such as~\texttt{Int} or
        \texttt{Ptr}) exactly once, just evaluate it;
        % \item
        to consume a function value exactly once, apply it to one argument,
        and consume its result exactly once;
        % \item
        To consume a value of an algebraic datatype exactly once,
        pattern-match on it, and consume all its linear components exactly once.
        % For example, a linear pair (equivalent to $\tensor$) is consumed exactly
        % once if pattern-matched on \emph{and} both the first and second element are
        % consumed once.
%\end{itemize}
Thus, we can consume a linear resource by fully evaluating it, through case
expressions.

@@

We start with an
example of a program that constructs a tuple from linear resources then pattern
matches on it, then uses both linearly-bound variables from the tuple pattern
match. This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f6 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f6 x y use = case (x,y) of { (a,b) -> use a b }
\end{lstlisting}
\end{working}
What might be more surprising is that a similar program which discards the
pattern variables and instead uses the resources in the scrutinee is 
\emph{semantically} linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f7 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f7 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
We justify that \incode{f7} is linear by appealing to the fact that 
the tuple being scrutinized is already in WHNF and so evaluating it will
not consume either $x$ nor $y$. Even if the tuple was constructed with
two expressions using $x$ and $y$ respectively, no computation would happen
since we are not using $a$ or $b$ (thereby never forcing the arguments
of the tuple). However, if we did use $a$ in the case body, then $x$ would be unavailable:
\begin{noway}
\begin{lstlisting}
f8 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f8 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is the same behind let
bindings being mutually exclusive to the resources that define them.
By forcing the pattern variable we run the computations
defined in terms of the linear variables used for that constructor
argument. Otherwise, if we do not use the pattern variables, then we
do not run the computation and thus no resources are consumed.

A third option is to use the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{lstlisting}
f9 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f9 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{lstlisting}
\end{notyet}
Again, $z$ is mutually exclusive with $a,b$ and with $x,y$, but at least one of
the three must occur to ensure the linear resources are consumed. In this
example, we can think that using $a$ entails using the resource $x$, $b$ the
resource $y$, and the case binder $z$ entails using both $a$ and $b$.
%
Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
evaluating it to WHNF will require performing computation and thus consume linear
resources:
\begin{notyet}
\begin{lstlisting}
f10 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
f10 x y use = case use x y of z { (a,b) -> z }
\end{lstlisting}
\end{notyet}
Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
alternatives, but we \emph{must} still use either the case binder $z$ or the linear
pattern variables $a,b$. %  e.g. it would be quite disastrous if any of the following typechecked:
% \begin{noway}
% \begin{lstlisting}
% doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
% doubleFree x free = case free x of z { Result v -> free x }
% \end{lstlisting}
% \end{noway}
% % Remover
% \begin{noway}
% \begin{lstlisting}
% leakPointer :: Ptr -> ()
% leakPointer x = case id x of z { _ -> () }
% \end{lstlisting}
% \end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
The result of evaluating the scrutinee must be consumed exactly once to guarantee
that the resources used in the scrutinee are fully consumed. %, or
                                %risk them being only ``almost''
                                %consumed.
For instance, take
\incode{use} in \incode{f10} to be \incode{(,)}: it is not sufficient for \incode{use x y} to be
evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
considered to be fully consumed after the scrutinee were evaluated in a case
expression, we could simply ignore the pattern variables, potentially
discarding linear resources. In
short, if the scrutinee is not in WHNF we must either consume the case binder
or the linear components of the pattern.

We now focus on pattern matching on constructors without
any linear components. If the constructor has no linear fields, it means the
result can be consumed unrestrictedly and, therefore, all linear resources used
in the computation have been fully consumed.
%
Consequently, in a branch of a constructor without linear fields we know the
result of evaluating the scrutinee to be unrestricted, so we can use the case
binder unrestrictedly and refer to it zero or more times. For example,
the program in Figure~\ref{fig:nolin} is semantically linear.

\begin{figure}[t]
 \begin{minipage}{0.5\textwidth}
\begin{notyet}
\begin{lstlisting}
f11 :: () *' $\lolli$ '* ()
f11 x = case x of z { () -> z <> z }
\end{lstlisting}
\end{notyet}
\vspace{-0.5cm}
\caption{No Linear Fields\label{fig:nolin}}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
\begin{limitation}
\begin{lstlisting}
f12 :: a *' $\lolli$ '* a
f12 x = case K1 x of z { K2 -> x; K1 a -> x }
\end{lstlisting}
\end{limitation}
\vspace{-0.5cm}
\caption{Absurd Branches\label{fig:absurd}}
\end{minipage}
\end{figure}

A second example of an unrestricted pattern, where \incode{K2} has no fields and \incode{K1}
has one linear field, is shown in Figure~\ref{fig:absurd}. Seemingly, the resource
$x$ must have been fully consumed to take the \incode{K2} branch, but because the
scrutinee is known to be \incode{K1}, the \incode{K2} branch is \emph{absurd}, so, in reality
any resource could be freely used in that branch, and the example is
\emph{semantically} linear (despite not being seen as so by our system).
%
%
%
This particular example might seem unrealistic due to the pattern
match on a known constructor, but we note that during the
transformations programs undergo in an optimising compiler, many
programs such as this naturally occur (e.g.,~if the definition of a
function is inlined in the scrutinee).
%
% However, the scrutinee before evaluation likely isn't in WHNF, thus scrutinee
% resources cannot directly occur in the case alternatives, since they might be
% potentially consumed when evaluating the scrutinee, the program leading up to
% this one wouldn't have been linear (even semantically).

>>>>

%\vspace{-.5cm}
\paragraph{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}
% Second, we look into recursive let bindings.
For the most part,
recursive let bindings behave as non-recursive let bindings, i.e. we must use them \emph{at
most once} because, when evaluated, the linear resources used in the binders
bodies are consumed. The defining property of a group of mutually recursive let
bindings is that the binders bound in that group can occur, without restrictions, in
the bodies of those same binders. % The same way that, in a let body, evaluating
% a binding that uses some resource exactly once consumes that resource once,
% using the binding in its own definition also entails using that resource once.
Consider the following program, that calls a recursive let-bound function
defined in terms of the linear resource \incode{x} and itself:
%
\begin{notyet}
\begin{lstlisting}
f3 :: Bool -> a *' $\lolli$ '* a
f3 bool x = let go b = case b of
                     True -> x
                     False -> go (not b)
                   in go bool
\end{lstlisting}
\end{notyet}
%
Function \incode{f3} is semantically linear because, iff it is consumed exactly once,
then \incode{x} is consumed exactly once. We can see this by case analysis on \incode{go}'s argument:
%\begin{itemize}%
%\item
when \incode{bool} is \incode{True}, we'll use the resource \incode{x};
%\item
 when \incode{bool} is \incode{False}, we recurse by calling \incode{go} on \incode{True}, which in turn will use the resource \incode{x}.
%\end{itemize}
In \incode{go}'s body, \incode{x} is used directly in one branch and indirectly in the
other (by recursing, which we know will result in using \incode{x} linearly).
% ROMES:TODO:!!
%\todo{this bit is quite hard to explain. It is some sort of cyclic
%argument -- we kind of assume go uses x linearly s.t. when go itself is used
%then we're using x linearly. recursion...}
%

@@

\paragraph{Case expressions}
Finally, we discuss case expressions,
which are the key ingredient that brings together the
semantic linearity insights developed thus far given that
\emph{case expressions drive evaluation}, and semantic linearity can only be
understood in terms of how expressions are evaluated.

Up until now, there
have been no examples in which linear resources are \emph{fully consumed} in
the bodies of linear functions. In other words, all example functions so far
return a value that has to be itself consumed exactly once to ensure the linear
argument is, in turn, consumed exactly once -- as opposed to functions whose
application simply needs to be evaluated to guarantee its linear argument is
consumed (e.g.,~functions that return an unrestricted value).
%
%The latter are of particular relevance because linearly-typed abstractions
%usually require such a function to provide newly-created linear resources to.
%
% For example, the entry point to the linear array API presented in Linear
% Haskell takes such a function as its second argument:
% \begin{working}
% \begin{lstlisting}
% newMArray :: Int -> (MArray a *' $\lolli$ '* Unrestricted b) *' $\lolli$ '* b
% \end{lstlisting}
% \end{working}
%(The second argument is a function that consumes |MArray a| linearly
%and returns an unrestricted result -- we don't need to consume said result
%exactly once to guarantee that |MArray a| is used linearly in the function
%body.)
%
% ROMES:TODO: Há qualquer coisa que não gosto nada! na coerência deste paragrafo
%
Case expressions enable us to consume resources and so write
functions that fully consume their linear arguments.
% CORTEI
To understand exactly how,
we turn to the definition of \emph{consuming} a resource from Linear
Haskell~\cite{cite:linearhaskell}:
%begin{itemize}
% \item
to consume a value of atomic base type (such as~\texttt{Int} or
        \texttt{Ptr}) exactly once, just evaluate it;
        % \item
        to consume a function value exactly once, apply it to one argument,
        and consume its result exactly once;
        % \item
        to consume a value of an algebraic datatype exactly once,
        pattern-match on it, and consume all its linear components exactly once.
        % For example, a linear pair (equivalent to $\tensor$) is consumed exactly
        % once if pattern-matched on \emph{and} both the first and second element are
        % consumed once.
%\end{itemize}
Thus, we can consume a linear resource by fully evaluating it, through case
expressions.

@@

We start with an
example of a program that constructs a tuple from linear resources then pattern
matches on it, then uses both linearly-bound variables from the tuple pattern
match. This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f6 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f6 x y use = case (x,y) of z { (a,b) -> use a b }
\end{lstlisting}
\end{working}
What might be more surprising is that a similar program which discards the
pattern variables and instead uses the resources in the scrutinee is 
\emph{semantically} linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f7 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f7 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
We justify that \incode{f7} is linear by appealing to the fact that 
the tuple being scrutinized is already in WHNF and so evaluating it will
not consume either $x$ nor $y$. Even if the tuple was constructed with
two expressions using $x$ and $y$ respectively, no computation would happen
since we are not using $a$ or $b$ (thereby never forcing the arguments
of the tuple). However, if we did use $a$ in the case body, then $x$ would be unavailable:
\begin{noway}
\begin{lstlisting}
f8 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f8 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is the same behind let
bindings being mutually exclusive to the resources that define them.
By forcing the pattern variable we run the computations
defined in terms of the linear variables used for that constructor
argument. Otherwise, if we do not use the pattern variables, then we
do not run the computation and thus no resources are consumed.

A third option is to use the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{lstlisting}
f9 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f9 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{lstlisting}
\end{notyet}
Again, $z$ is mutually exclusive with $a,b$ and with $x,y$, but at least one of
the three must occur to ensure the linear resources are consumed. In this
example, we can think that using $a$ entails using the resource $x$, $b$ the
resource $y$, and the case binder $z$ entails using both $a$ and $b$.
%
Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
evaluating it to WHNF will require performing computation and thus consume linear
resources:
\begin{notyet}
\begin{lstlisting}
f10 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
f10 x y use = case use x y of z { (a,b) -> z }
\end{lstlisting}
\end{notyet}
Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
alternatives, but we must still use either the case binder $z$ or the linear
pattern variables $a,b$. %  e.g. it would be quite disastrous if any of the following typechecked:
% \begin{noway}
% \begin{lstlisting}
% doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
% doubleFree x free = case free x of z { Result v -> free x }
% \end{lstlisting}
% \end{noway}
% % Remover
% \begin{noway}
% \begin{lstlisting}
% leakPointer :: Ptr -> ()
% leakPointer x = case id x of z { _ -> () }
% \end{lstlisting}
% \end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
The result of evaluating the scrutinee must be consumed exactly once to guarantee
that the resources used in the scrutinee are fully consumed. %, or
                                %risk them being only ``almost''
                                %consumed.
For instance, take
\incode{use} in \incode{f10} to be \incode{(,)}: it is not sufficient for \incode{use x y} to be
evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
considered to be fully consumed after the scrutinee were evaluated in a case
expression, we could simply ignore the pattern variables, potentially
discarding linear resources. In
short, if the scrutinee is not in WHNF we must either consume the case binder
or the linear components of the pattern.

We now focus on pattern matching on constructors without
any linear components. If the constructor has no linear fields, it means the
result can be consumed unrestrictedly and, therefore, all linear resources used
in the computation have been fully consumed.
%
Consequently, in a branch of a constructor without linear fields we know the
result of evaluating the scrutinee to be unrestricted, so we can use the case
binder unrestrictedly and refer to it zero or more times. For example,
the program in Figure~\ref{fig:nolin} is semantically linear.
%
\begin{figure}[t]
 \begin{minipage}{0.5\textwidth}
\begin{notyet}
\begin{lstlisting}
f11 :: () *' $\lolli$ '* ()
f11 x = case x of z { () -> z <> z }
\end{lstlisting}
\end{notyet}
\vspace{-0.5cm}
\caption{No Linear Fields\label{fig:nolin}}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
\begin{limitation}
\begin{lstlisting}
f12 :: a *' $\lolli$ '* a
f12 x = case K1 x of z { K2 -> x; K1 a -> x }
\end{lstlisting}
\end{limitation}
\vspace{-0.5cm}
\caption{Absurd Branches\label{fig:absurd}}
\end{minipage}
\end{figure}
%
A second example of an unrestricted pattern, where \incode{K2} has no fields and \incode{K1}
has one linear field, is shown in Figure~\ref{fig:absurd}. Seemingly, the resource
$x$ must have been fully consumed to take the \incode{K2} branch, but because the
scrutinee is known to be \incode{K1}, the \incode{K2} branch is \emph{absurd}, so, in reality
any resource could be freely used in that branch, and the example is
\emph{semantically} linear (despite not being seen as so by our system).

Many of these examples might seem unrealistic due to the pattern
match on a known constructor, but we note that, during the
transformations programs undergo in an optimising compiler, many
such programs naturally occur (e.g.,~if the definition of a
function is inlined in the scrutinee).
%
% However, the scrutinee before evaluation likely isn't in WHNF, thus scrutinee
% resources cannot directly occur in the case alternatives, since they might be
% potentially consumed when evaluating the scrutinee, the program leading up to
% this one wouldn't have been linear (even semantically).
